opXRD: Open Experimental Powder X-ray Diffraction Database

### 2024-05-14 chapters/00_abstract.tex
Pushed resolved comments

Powder X-ray diffraction (pXRD) experiments are a cornerstone for materials structure characterization.
Despite their widespread application, the analysis of pXRD diffractograms still presents a significant automation challenge and thus a bottleneck in high-throughput experimentation that remains to be solved in order to match increasing throughput of automated materials synthesis in (self-driving) labs.
In the current landscape of powder X-ray diffraction analysis, machine learning has emerged as a promising research direction to resolve this bottleneck by enabling automated powder diffraction analysis.
A notable difficulty in applying machine learning to this domain is the lack of sufficiently sized experimental datasets, which has relegated machine learning researchers to train primarily on simulated data. Since simulations largely fail to accurately reflect the experiment, the performance of models trained on only simulated data often lacks transferability to experimental data and thus fails to provide value in practice.
With the Open Experimental Powder X-Ray Diffraction Database (opXRD), we aim to remedy this by providing an openly available and easily accessible dataset of partially labeled experimental powder diffractograms, providing machine learning researchers with a large quantity of real experimental diffractograms collected from a broad range of samples.
We provide almost barrier-free software tools that allow experimental researchers to find their data and make it accessible - in virtually any widely used format.
We collected XXX labeled and YYY unlabeled diffractograms from a wide spectrum of materials classes, which establishes the first version of the opXRD database.
We hope that this ongoing effort can guide machine learning research toward fully automated analysis of pXRD data and thus enable future self-driving materials labs.

>>>>

Powder X-ray diffraction (pXRD) experiments are a cornerstone for materials structure characterization.
Despite their widespread application, the analysis of pXRD diffractograms still presents a significant challenge to automation and thus a bottleneck in high-throughput experimentation automated materials discovery in self-driving labs.
Machine learning has emerged as a promising research direction to resolve this bottleneck by enabling automated powder diffraction data analysis.
A notable difficulty in applying machine learning to this domain is the lack of sufficiently sized experimental datasets, which has relegated machine learning researchers to train primarily on simulated data. Since simulations largely fail to accurately reflect the experiment, the performance of models trained on only simulated data often lacks transferability to experimental data and thus fails to provide value in practice.
With the Open Experimental Powder X-Ray Diffraction Database (opXRD), we aim to remedy this by providing an openly available and easily accessible dataset of partially labeled experimental powder diffractograms, providing machine learning researchers with a large quantity of real experimental diffractograms collected from a broad range of samples.
We provide almost barrier-free software tools that allow experimental researchers to find their data and make it accessible - in virtually any widely used format.
We collected XXX labeled and YYY unlabeled diffractograms from a wide spectrum of materials classes, which establishes the first version of the opXRD database.
We hope that this ongoing effort can guide machine learning research toward fully automated analysis of pXRD data and thus enable future self-driving materials labs.

### 2024-05-14 chapters/01_introduction.tex
Pushed resolved comments

% On Rietveld refinement
After fabricating a new material, a number of analysis techniques can be used to characterize the sample.
The most important method to determine the crystal structure of a new material is powder X-ray diffraction. 
%In the case of multi-phase samples, this includes identifying the phases present in the sample and quantifying their fractions.
To analyze pXRD data, an initial guess is optimized using Rietveld refinement. In this process, an initial crystal structure, including unit cell parameters, provided by the operator of the analysis software, is fitted to the observed diffractogram in an iterative optimization process based on simulated forward prediction of expected pXRD diffractograms given the current crystal structure.
%The unit cell parameters of the crystal structure are adjusted to minimize the difference between the simulated diffraction pattern and the pattern observed in the experiment. 
As Rietveld refinement is a local optimization method, the result of the refinement procedure is generally only as good as the initial guess it was provided with.
Therefore, it strongly relies on the experience of the operator to provide an accurate initial guess of the crystal structure.

Employing manually operated Rietveld refinement, including determining a good initial structure for the optimization, is not without issues.
It is certainly not scalable to the degree required to keep up with advances in throughput and efficiency in other steps of the experimentation pipeline.
Furthermore, initial guesses for Rietveld refinement are often obtained by matching the observed diffractogram to patterns from a database.
%But beyond the lack of scalability of this workflow we also wish to highlight the following issue:
%Providing the analysis software with an initial guess of the structure requires manually or algorithmically
%consulting a reference library of crystal structures with corresponding simulated or sometimes real diffractograms.
%From that reference library, a crystal structure must be identified that is likely to be similar
%to that of the investigated sample which can then serve as the initial guess provided to the analysis software.
%Since the number of chemically viable crystal structures is infinite, any such library will always be incomplete.
However, in the case of novel observed structure types, simply matching to a database of known structures will not yield good results. Therefore, more complex crystal structure solution approaches need to be applied to find suitable crystal structure guesses.
%The best that can be done to remedy this is to continually update the library both on the side
%of the distributor and the user, which is an undesirable overhead.
%Especially for novel material discovery, it may be impossible to find a sufficiently accurate match in the reference
%library to apply Rietveld Refinement successfully.

@@

At the moment, the number of openly available datasets for experimental powder data is very limited.
The main sources for experimental patterns are XXX, YYY, and ZZZ.
In this publication, we provide a new dataset that collects a broad range of patterns from experiment, exceeding the currently available amount of openly accessible experimental patterns by XX.
The dataset has been curated by collecting the accumulated powder data from multiple large research groups and institutions with high-throughput XRD facilities.

% How to contribute? This aspect deserves a separate paragraph
We further provide an online platform that can be used to easily submit new pXRD data to the dataset, which will be used to extend the existing dataset in the future (Figure~\ref{fig:overview}).
TODO: Make this longer?

Collecting a labeled dataset of raw experimental powder X-ray diffraction data of sufficient size to directly train a model that predicts crystal structure information is still out of reach.
The most advanced ML models currently are trained on approximately x simulated diffractograms, which is x orders of magnitude higher than the initial state of the opXRD database presented here.
However, our dataset still holds significant value in practice. First of all, the dataset can be used to better test and benchmark existing and new automated analysis approaches, which yields a better comparison than when using the very limited currently available datasets, such as the commonly used RRUFF mineral dataset.

Furthermore, we hope that our dataset can enable machine learning researchers to evaluate how closely their simulations
represent actual experimental data, identify what features are missing, and modify their simulation algorithms accordingly to arrive at a simulation that mirrors how data appears in reality.
One can bring the simulated patterns closer to experiment by improving the simulation itself by properly modeling more of the observed physical effects such as preferred orientation, peak shapes resulting from different sources, etc.
Additionally, some of the observed imperfections, such as background and noise, can be modeled using machine learning approaches.
This is similar to the concept of style transfer\cite{Gatys2016,Ganin2015} found in the domain of machine learning applied to images.
Here, the objective is to obtain a classification or regression model when only labeled images of one style (simulated data) and unlabeled or partially labeled images of another style (experimental data) are available.
The broad range of available experimental samples contained in our dataset makes it possible to apply these state-of-the-art ML approaches to the domain of powder XRD analysis.
We hope that this will yield more advanced automated ML approaches in the future, which can readily be applied in high-throughput experimentation pipelines.

@@

% TODOS:
% - This paper talks about the importance of sharing raw powder x-ray diffraction data. I'd definitely cite it ->  \cite{Aranda2018}
% - The image diff.png shows the difference between a powder diffractogram calculated from single x-ray diffraction data
% and a real powder diffractogram of the same material. The difference is enormous. I'd include it to highlight
% that simulating powder diffractino patterns from single crystal data only goes so far
%https://www.sciencedirect.com/topics/biochemistry-genetics-and-molecular-biology/rietveld-refinement

>>>>

% On Rietveld refinement
After fabricating a new material, a number of analysis techniques can be used to characterize the sample.
The most important method to determine the crystal structure of a new material is powder X-ray diffraction. 
%In the case of multi-phase samples, this includes identifying the phases present in the sample and quantifying their fractions.
To analyze pXRD data, an initial guess is optimized using Rietveld refinement. In this process, an initial crystal structure including unit cell parameters is provided by the operator of the analysis software and fitted to the observed diffractogram in an iterative optimization process based on simulated forward prediction of expected pXRD diffractograms given the current crystal structure.
%The unit cell parameters of the crystal structure are adjusted to minimize the difference between the simulated diffraction pattern and the pattern observed in the experiment. 
As Rietveld refinement is a local optimization method, the result of the refinement procedure is generally only as good as the initial guess it was provided with.
Therefore, it strongly relies on the experience of the operator to provide an accurate initial guess of the crystal structure.

Employing manually operated Rietveld refinement, including determining a good initial structure for the optimization, is not without issues.
It is certainly not scalable to the degree required to keep up with advances in throughput and efficiency in other steps of the experimentation pipeline.
Automation of the Rietveld refinement requires the automation of finding initial guess structures, which is highly non-trivial, and the automation of the iterative optimization procedure, which can be done using black-box optimization algorithms.
Initial guesses for Rietveld refinement are often obtained by matching the observed diffractogram to patterns from a database.
%But beyond the lack of scalability of this workflow we also wish to highlight the following issue:
%Providing the analysis software with an initial guess of the structure requires manually or algorithmically
%consulting a reference library of crystal structures with corresponding simulated or sometimes real diffractograms.
%From that reference library, a crystal structure must be identified that is likely to be similar
%to that of the investigated sample which can then serve as the initial guess provided to the analysis software.
%Since the number of chemically viable crystal structures is infinite, any such library will always be incomplete.
However, in the case of novel observed structure types, simply matching to a database of known structures will not yield good results. Therefore, more complex crystal structure solution approaches need to be applied to find suitable crystal structure guesses.
%The best that can be done to remedy this is to continually update the library both on the side
%of the distributor and the user, which is an undesirable overhead.
%Especially for novel material discovery, it may be impossible to find a sufficiently accurate match in the reference
%library to apply Rietveld Refinement successfully.

@@

At the moment, the number of openly available datasets for experimental powder data is very limited.
The main sources for experimental patterns are XXX, YYY, and ZZZ.
Further commercial datasets such as KKK and LLL exist, but license fees and restrictions in their use to train publishable models or even publish the training data limit their usability.
Collecting a labeled dataset of raw experimental powder X-ray diffraction data of sufficient size to directly train a model that predicts crystal structure information is likely out of reach.
The most advanced ML models currently are trained on approximately x simulated diffractograms, which is x orders of magnitude higher than the largest currently curated experimental dataset.

However, even small labeled and also unlabeled experimental diffractograms hold significant value in practice. First of all, labeled data can be used to better test and benchmark existing and new automated analysis approaches, which yields a better comparison than when using the very limited currently available datasets, such as the commonly used RRUFF mineral dataset.

Furthermore, even completely unlabeled data can enable machine learning researchers to evaluate how closely their simulations represent actual experimental data, identify what features are missing, and modify their simulation algorithms accordingly to arrive at a simulation that mirrors how data appears in reality.
One can bring the simulated patterns closer to experiment by improving the simulation itself by properly modeling more of the observed physical effects such as preferred orientation, peak shapes resulting from different sources, etc.
Additionally, some of the observed imperfections, such as background and noise, can be modeled using machine learning approaches.
This is similar to the concept of style transfer\cite{Gatys2016,Ganin2015} found in the domain of machine learning applied to images.
Here, the objective is to obtain a classification or regression model when only labeled images of one style (simulated data) and unlabeled or partially labeled images of another style (experimental data) are available.

In this publication, we provide a new open powder X-ray diffraction (opXRD) dataset that collects a broad range of patterns from experiment, exceeding the currently available amount of openly accessible experimental patterns by XX.
Our opXRD dataset has been curated by collecting the accumulated powder data from multiple large research groups and institutions with high-throughput XRD facilities.

% How to contribute? This aspect deserves a separate paragraph
The opXRD database is planned as a growing, community-driven initiative. The dataset we present here is the first version, but we hope to further increase the dataset size though active engagement with the pXRD community. In order to xxx. We further provide an online platform that can be used to easily submit new pXRD data to the dataset, which will be used to extend the existing dataset in the future (Figure~\ref{fig:overview}).
TODO: Make this longer?

The broad range of available experimental samples contained in our dataset makes it possible to apply these state-of-the-art ML approaches to the domain of powder XRD analysis.
We hope that this will yield more advanced automated ML approaches in the future, which can readily be applied in high-throughput experimentation pipelines.

@@

% - The image diff.png shows the difference between a powder diffractogram calculated from single x-ray diffraction data
% and a real powder diffractogram of the same material. The difference is enormous. I'd include it to highlight
% that simulating powder diffractino patterns from single crystal data only goes so far
%https://www.sciencedirect.com/topics/biochemistry-genetics-and-molecular-biology/rietveld-refinement
